Comparing Machine Learning algorithms for classifying cognitive states from fMRI data

### 2012-11-12 article/article.tex
review

The dataset used for our experiments was originally collected by Marcel Just and colleagues from Carnegie Mellon University's Center for Cognitive Brain Imaging and published as the StarPlus fMRI dataset. This set consists of fMRI data for 6 subjects that was obtained while having each subject participate in a number of trials.

In some of these trials (baseline trials), no stimulus was presented and participants were instructed to remain inert. In other trials, subjects were presented with a sequence of two stimuli on a computer screen. The first stimulus was either a picture or a sentence lasting 4 seconds, followed by a blank screen of 4 seconds. After the blank screen, the opposite type of stimulus was presented for 4 seconds, followed by a blank screen of 15 seconds, during which the subject would indicate by pressing a button whether the sentence correctly described the picture. During each trial, approximately 54 fMRI scans have been taken with an interval of 500ms. In total, 14 trials were baseline trials, while half of the remaining trials had a picture as first stimulus and the other half a sentence.  
The entire dataset consists of voxel data and meta-data providing information about the dataset. Included in this information is not only the trial and scan each voxel belongs to, but also the type of the trial (picture first or sentence first). This makes the dataset suitable for classification using supervised learning.

In total, we have data available for approximately 4800 voxels per scan, 54 scans per trial, 40 trials per subject (excluding baseline trials) and 6 subjects. Voxels have been grouped into 25 ROIs (see \textit{Feature selection}).

\subsection{Preprocessing}
\label{sec:preprocessing}
The dataset contains activation values for each voxels with respect to a baseline activation value for a subject, measured before performing the task. However, each subject may respond differently in terms of strength of voxel activation, making it harder to classify across subjects. To counter this, and make optimal use of the information carried by each voxel, we have created normalized copies of the dataset to use as input for the algorithms. The effect of this preprocessing step depends on assumptions made about the form of our data, which are not available. Instead, we had to verify the effect by looking at the resulting classification results (cf.~Table~\ref{tab:results}).

>>>>

The dataset used for our experiments was originally collected by Marcel Just et al. from Carnegie Mellon University's Center for Cognitive Brain Imaging and published as the StarPlus fMRI dataset. This set consists of fMRI data for 6 subjects that was obtained while having each subject participate in a number of trials.

In some of these trials (baseline trials), no stimulus was presented and participants were instructed to remain inert. In other trials, subjects were presented with a sequence of two stimuli on a computer screen. The first stimulus was either a picture or a sentence lasting 4 seconds, followed by a blank screen of 4 seconds. After the blank screen, the opposite type of stimulus was presented for 4 seconds, followed by a blank screen of 15 seconds, during which the subject would indicate by pressing a button whether the sentence correctly described the picture. During each trial, approximately 54 fMRI scans have been taken with an interval of 500ms. In total, 14 trials were baseline trials, while half of the remaining trials had a picture as first stimulus and the other half a sentence.

The entire dataset consists of voxel data and meta-data providing information about the dataset. Included in this information is not only the trial and scan each voxel belongs to, but also the type of the trial (picture first or sentence first). This makes the dataset suitable for classification using supervised learning.

In total, we have data available for approximately 4800 voxels per scan, 54 scans per trial, 40 trials per subject (excluding baseline trials) and 6 subjects. Voxels have been grouped into 25 ROIs (cf.~Section~\ref{sec:dimensionalityReduction}).

\subsection{Preprocessing}
\label{sec:preprocessing}

The dataset contains activation values for each voxels with respect to a baseline activation value for a subject, measured before performing the task. However, each subject may respond differently in terms of strength of voxel activation, making it harder to classify across subjects. To counter this, and make optimal use of the information carried by each voxel, we have created normalized copies of the dataset to use as input for the algorithms. The effect of this preprocessing step depends on assumptions made about the form of our data, which are not available. Instead, we had to verify the effect by looking at the resulting classification results (cf.~Table~\ref{tab:results}).
